---
title: "Shift Happens"
author: "Felipe MENDEZ and Benjamin RENARD (INRAE)"
date: "February 2026"
output:
  github_document:
    toc: yes
subtitle: Documentation
bibliography: inst/RatingShiftHappens.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  comment = "#>",
  fig.path = "man/readme/README-",
  out.width = "100%"
)
```

# Introduction

`ShiftHappens` is an `R` package for detecting, visualizing and estimating shifts, with a specific focus on hydrology and hydrometry applications. The package is derived from previous works, in particular:

1.  [BayDERS](https://github.com/MatteoDarienzo/BayDERS) developed by [@darienzoDetectionEstimationStagedischarge2021](https://theses.hal.science/tel-03211343).
2.  [RatingShiftHappens](https://github.com/Felipemendezrios/RatingShiftHappens), a previous version of this package.

```{r  message=FALSE}
# devtools::install_github("benRenard/ShiftHappens") # Install before first use
library(ShiftHappens)
```

# Segmentation of a series of observations

The most common segmentation task is to look for changes in a time series.

## Motivating dataset

The dataset `RhoneRiverAMAX` coming with the package contains annual maximum stages (`H`, $m$) and discharges (`Q`, $m^3.s^{-1}$) for the Rhône River at Beaucaire, France, along with the associated uncertainties expressed as standard deviations (`uH` and `uQ`). A shift is visually apparent in the stage series around 1970. For more details on this dataset, see [@lucas2023](https://hal.inrae.fr/hal-04170646).

```{r Beaucaire}
 plot(RhoneRiverAMAX$Date,RhoneRiverAMAX$H)
```

## Basic usage

To segment a series of observations without knowing the number of segments, we recommend starting with the function `Segmentation_Recursive`. With its default arguments, this function attempts to split the series into two segments using the method described in [@gombay1994](https://www.sciencedirect.com/science/article/pii/0304414994901546), then if successful tries to further split each obtained segment into two sub-segments, and so one until all segments cannot be further split. The function returns an object (see `?recursiveSegmentation`) that can be plotted (see `?plot.recursiveSegmentation`).

```{r Segmentation_Recursive}
sg=Segmentation_Recursive(time=RhoneRiverAMAX$Date,
                         obs=RhoneRiverAMAX$H,
                         u=RhoneRiverAMAX$uH)
sg$shifts # estimated shifts
plot(sg)
```

For an illustration with more detected segments, consider segmenting the series of discharge uncertainties `uQ`: the many detected changes reflect changes in the measurement process along the years.

```{r Segmentation_Recursive2}
sg=Segmentation_Recursive(time=RhoneRiverAMAX$Date,
                          obs=RhoneRiverAMAX$uQ)
sg$shifts
plot(sg)
```

It is also possible to visualize the recursion tree:

```{r Tree}
plot(sg$tree)
```

## Advanced usages

If the number of segments is known, the function `Segmentation_Engine` can be used. It returns an object (see `?simpleSegmentation`) that can be plotted. Note that if the desired number of segments is larger than two, the aforementioned method of @gombay1994 is replaced by the bayesian method described by [@darienzoDetectionStageDischarge2021](https://doi.org/10.1029/2020WR028607). The package [RBaM](https://github.com/BaM-tools/RBaM) needs to be installed in this case.

```{r Segmentation_Engine}
sg=Segmentation_Engine(time=RhoneRiverAMAX$Date,
                       obs=RhoneRiverAMAX$uQ,
                       nS=4)
plot(sg)
```

When the number of segments is unknown, the function `Segmentation` will try all possible numbers up to `nSmax`, and automatically select the most appropriate.

```{r Segmentation}
sg=Segmentation(time=RhoneRiverAMAX$Date,
                obs=RhoneRiverAMAX$H,
                nSmax=4)
plot(sg)
```

# Segmentation based on recursive modeling

In some cases, the objective is not to assess whether the properties of a variable $x$ has changed, but rather to assess whether *the relation* between two variables $x$ and $y$ has changed. A possible approach for doing so is to model the relation between $x$ and $y$ and to look for changes in the residuals of the model (i.e. observed values of $y$ minus predicted values).

## Motivating dataset

The dataset `ArdecheRiverGaugings` coming with the package contains joint measurements of stage (`H`, $m$) and discharge (`Q`, $m^3.s^{-1}$) for the Ardèche River at Meyras station, France, along with the associated discharge uncertainties expressed as standard deviations (`uQ`). The figure suggests the existence of at least two distinct relations. For more details on this dataset, see [@mansanarezShiftHappensAdjusting2019](https://doi.org/10.1029/2018WR023389).

```{r Meyras}
plot(ArdecheRiverGaugings$H,ArdecheRiverGaugings$Q)
```

## Recursive linear regression

The function `Segmentation_RecursiveModeling` fits a model between $x$ and $y$ and then segment the residuals. If several segments are detected, the model is fitted again on each obtained segment and the segmentation is re-applied, and so one until all segments cannot be further split. By default the model used is a simple linear regression between $x$ and $y$. 
The function returns an object (see `?recursiveModeling`) that can be plotted (see `?plot.recursiveModeling`).

```{r Recursive_Linear}
sg=Segmentation_RecursiveModeling(x=ArdecheRiverGaugings$H,
                                  y=ArdecheRiverGaugings$Q,
                                  uY=ArdecheRiverGaugings$uQ,
                                  time=ArdecheRiverGaugings$Date)
plot(sg)
```

It is possible to replace the upper panel showing the relation between $x$ and $y$ by a time series of $y$ (or $x$).

```{r Recursive_Linear2}
plot(sg,dataPlotType='ty')
```

## Recursive application of the BaRatin rating curve method

The linear regression model illustrated previously is not a very good model for the relation between stage $H$ and discharge $Q$. A [rating curve model](https://hal.science/hal-00934237/) such as the one described by @lecozCombiningHydraulicKnowledge2014 is a much better choice. The function `Segmentation_RecursiveModeling` has an argument `Fit_funk` that allows passing a function fitting any other model. Here the [BaRatin model](https://baratin-tools.github.io/en/), specifically designed for rating curves, is used. A detailed description of the BaRatin method can be found [here](https://baratin-tools.github.io/en/doc/baratinage/), with a specific explanation on how to use BaRatin in `R` being given [here](https://baratin-tools.github.io/en/doc/case/using-rbam/).

```{r Recursive_BaRatin}
# Define control matrix: columns are controls, rows are stage ranges.
# See https://baratin-tools.github.io/en/doc/topics/hydraulic-analysis/
controlMatrix=rbind(c(1,0,0),c(0,1,0),c(0,1,1))
# Define parameters of the model
k1=RBaM::parameter(name='k1',init=-0.6,prior.dist='Gaussian',prior.par=c(-0.6,0.5))
a1=RBaM::parameter(name='a1',init=exp(2.65),prior.dist='LogNormal',prior.par=c(2.65,0.35))
c1=RBaM::parameter(name='c1',init=1.5,prior.dist='Gaussian',prior.par=c(1.5,0.025))
k2=RBaM::parameter(name='k2',init=0,prior.dist='Gaussian',prior.par=c(0,1))
a2=RBaM::parameter(name='a2',init=exp(3.28),prior.dist='LogNormal',prior.par=c(3.28,0.33))
c2=RBaM::parameter(name='c2',init=1.67,prior.dist='Gaussian',prior.par=c(1.67,0.025))
k3=RBaM::parameter(name='k3',init=1.2,prior.dist='Gaussian',prior.par=c(1.2,0.2))
a3=RBaM::parameter(name='a3',init=exp(3.46),prior.dist='LogNormal',prior.par=c(3.46,0.38))
c3=RBaM::parameter(name='c3',init=1.67,prior.dist='Gaussian',prior.par=c(1.67,0.025))
priors=list(k1,a1,c1,k2,a2,c2,k3,a3,c3)
# Note: possible to use wrapper function ?Segmentation_BaRatin instead of the function below
sg=Segmentation_RecursiveModeling(x=ArdecheRiverGaugings$H,
                                  y=ArdecheRiverGaugings$Q,
                                  uY=ArdecheRiverGaugings$uQ,
                                  time=ArdecheRiverGaugings$Date,
                                  Fit_funk=Fit_BaRatin,
                                  controlMatrix=controlMatrix,priors=priors)
plot(sg)
```

Note that each individual plot created by the package is a ggplot that can be customized. For instance:

```{r Recursive_BaRatin2}
plot(sg,type='data')+ggplot2::labs(x='Stage',y='Discharge')+
  ggplot2::scale_y_continuous(trans='log')
```

# Segmentation based on stage recessions

This method is specific to the hydrometry context, and aims at detecting shifts using the stage record, measured at any hydrometric station. The basic idea is that the water level tends to the riverbed level at very low flows. It should therefore be possible to detect changes by looking at the lowest water levels reached during long recession periods.

## Motivating dataset

The dataset `ArdecheRiverStage` coming with the package contains the stage time series (`H`, $m$) for the Ardèche River at Meyras station, France. There seems to be a downward shift around 2008 in the lowest stage values. For more details on this dataset, see [@darienzoDetectionEstimationStagedischarge2021](https://theses.hal.science/tel-03211343).

```{r MeyrasStage}
plot(ArdecheRiverStage$Date,ArdecheRiverStage$H,type='l')
```

## Recession extraction

The first step is to extract recession events using the function `Extract_Recessions`. Many options are available to customize the definition of recession events, see `?Extract_Recessions`. The function returns an object (see `?extractedRecessions`) that can be plotted in various ways (see `?plot.extractedRecessions`). Here for instance the extracted recession events are shown as a function of the calendar date (left) or the within-recession time (i.e. starting at the beginning of the recession, right).

```{r extractRec}
rec=Extract_Recessions(time=ArdecheRiverStage$Date,H=ArdecheRiverStage$H,dMin=30)
patchwork::wrap_plots(plot(rec),plot(rec,type='th'))
```

## Recession segmentation

The next step is to apply a segmentation procedure to a well-chosen property of the extracted recession events. The simplest way to start is to get the lowest value of each recession (using utility function `getRecessionMin`) and to segment them.

```{r recSegment}
lows=getRecessionMin(rec)
sg=Segmentation_Recursive(obs=lows$H,time=lows$date)
plot(sg)
```

A more advanced approach is to model the extracted recessions and to segment the parameter of this model that controls the asymptotic low stage. Details on this procedure can be found in [@darienzoDetectionEstimationStagedischarge2021](https://theses.hal.science/tel-03211343) (Chapter 3). First, the function `Fit_Recessions` can be used to fit the recession model.

```{r recModel}
getRecessionEquations() # View available recession equations
f=Fit_Recessions(rec,equation='M7',
                 # The line below aims at speeding up computations, 
                 # but it is advised to make definitive runs 
                 # with the default MCMC options
                 mcmc_options=RBaM::mcmcOptions(nAdapt=25,nCycles=20))
plot(f)
```

Second, the parameters controling the recession asymptotic stages can be extracted and segmented. Note that the wrapper function `Segmentation_Recessions` can be used to perform the whole model-then-segment analysis in one go. 

```{r recModel2}
betas=f$parameters[1:max(rec$index),]
sg=Segmentation_Recursive(obs=betas$value,u=betas$u)
plot(sg)
```

# References
