---
title: "Shift Happens"
author: "Felipe MENDEZ and Benjamin RENARD (INRAE)"
date: "February 2026"
output:
  github_document:
    toc: yes
subtitle: Documentation
bibliography: inst/RatingShiftHappens.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  comment = "#>",
  fig.path = "man/readme/README-",
  out.width = "100%"
)
```

# Introduction

`ShiftHappens` is an `R` package for detecting, visualizing and estimating shifts, with a specific focus on hydrology and hydrometry applications. The package is derived from previous works, in particular:

1.  [BayDERS](https://github.com/MatteoDarienzo/BayDERS) developed by [\@darienzoDetectionEstimationStagedischarge2021](https://theses.hal.science/tel-03211343).
2.  [RatingShiftHappens](https://github.com/Felipemendezrios/RatingShiftHappens), a previous version of this package.

```{r  message=FALSE}
# devtools::install_github("benRenard/ShiftHappens") # Install before first use
library(ShiftHappens)
```

# Segmentation of a series of observations

## Motivating dataset

The dataset `RhoneRiverAMAX` coming with the package contains annual maximum stages (`H`, $m$) and discharges (`Q`, $m^3.s^{-1}$) for the Rhone River at Beaucaire, France, along with the associated uncertainties expressed as standard deviations (`uH` and `uQ`). For more details on this dataset, see [\@lucas2023](https://hal.inrae.fr/hal-04170646).

```{r Beaucaire}
 plot(RhoneRiverAMAX$Date,RhoneRiverAMAX$H)
```

## Basic usage

To segment a series of observations without knowing the number of segments, we recommend starting with the function `Segmentation_Recursive`. With its default arguments, this function attempts to split the series into two segments using the method described in [\@gombay1994](https://www.sciencedirect.com/science/article/pii/0304414994901546), then if successful tries to further split each obtained segment into two sub-segments, and so one until all segments cannot be further split. The function returns an object (see `?recursiveSegmentation`) that can be plotted (see `?plot.recursiveSegmentation`).

```{r Segmentation_Recursive}
sg=Segmentation_Recursive(time=RhoneRiverAMAX$Date,
                         obs=RhoneRiverAMAX$H,
                         u=RhoneRiverAMAX$uH)
plot(sg)
```

For an illustration with more detected segments, consider segmenting the series of discharge uncertainties `uQ`: the many detected changes reflect changes in the measurement process along the years.

```{r Segmentation_Recursive2}
sg=Segmentation_Recursive(time=RhoneRiverAMAX$Date,
                          obs=RhoneRiverAMAX$uQ)
plot(sg)
```

It is also possible to visualize the recursion tree:

```{r Tree}
plot(sg$tree)
```

## Advanced usages

If the number of segments is known, the function `Segmentation_Engine` can be used. It returns an object (see `?simpleSegmentation`) that can be plotted. Note that if the desired number of segments is larger than two, the aforementioned method of @gombay1994 is replaced by the bayesian method described by [\@darienzoDetectionStageDischarge2021](https://doi.org/10.1029/2020WR028607). You need to install the package [RBaM](https://github.com/BaM-tools/RBaM) in this case.

```{r Segmentation_Engine}
sg=Segmentation_Engine(time=RhoneRiverAMAX$Date,
                       obs=RhoneRiverAMAX$uQ,
                       nS=4)
plot(sg)
```

When the number of segments is unknown, the function `Segmentation` will try all possible numbers up to `nSmax`, and automatically select the most appropriate.

```{r Segmentation}
sg=Segmentation(time=RhoneRiverAMAX$Date,
                obs=RhoneRiverAMAX$H,
                nSmax=4)
plot(sg)
```

# Segmentation based on recursive modeling

## Motivating dataset

## Recursive linear regression

## Recursive application of the BaRatin rating curve method

# Segmentation based on stage recessions

## Motivating dataset

## Recession extraction

## Recession segmentation

# References
